\paperTitle

### 2014-04-28 doc/icml2014/abstract.tex

Recent work on the method of moments provide consistent estimates
  for latent-variable models, avoiding local optima issues,
  but these methods are applicable only to certain types of graphical models.
  On the other hand, pure likelihood objectives are difficult to optimize.
In this work, we show that using the method of moments in conjunction
  with composite marginal likelihood
  yields consistent parameter estimates for a much broader
  class of discrete directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we use tensor factorization to reveal partial information
about the hidden variables.  This renders the otherwise non-convex
likelihood objective convex, and EM can then be used to attain the global optimum.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
%Our approach gracefully extends to models outside our class by
  %incorporating the partial information via posterior regulraization.
\end{abstract}

>>>>

Recent work on the method of moments provide consistent estimates
  for latent-variable models, avoiding local optima issues,
  but are applicable only to certain types of graphical models.
  On the other hand, pure likelihood objectives are difficult to optimize.
In this work, we show that using the method of moments in conjunction
  with composite marginal likelihood
  yields consistent parameter estimates for a much broader
  class of discrete directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we use tensor factorization to reveal partial information
about the hidden variables.  This renders the otherwise non-convex
likelihood objective convex, and EM can then be used to attain the global optimum.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
%Our approach gracefully extends to models outside our class by
  %incorporating the partial information via posterior regulraization.
\end{abstract}

### 2014-04-28 doc/icml2014/directed.tex

\paragraph{Estimating $\pi$}
The moments of $x^a_1$, $\mO_1 \eqdef \Pr(x^a_1)$ are directly related to
  $\pi$ by $\mO_1 = O \pi$. 
By \assumptionref{full-rank}, $O$ has full column rank and thus can be
  inverted to recover $\pi$: $\pi = \pinv O \mO_1$.

\paragraph{Estimating $T$}
Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$, in terms of the hidden marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$:
\begin{align*}
\mO_{12} = O \mH_{12} O^\top \quad\Rightarrow\quad
  \mH_{12} = \pinv O \mO_{12} \pinvt O.
\end{align*}
$T$ can be recovered from the $\mH_{12}$ by suitable renormalization.

\paragraph{Estimating $V$}
Finally, we can estimate $V$ by renormalizing the hidden marginals
$\mH_{234} \eqdef \Pr(h_2, h_3, h_4)$ from the third-order moments
$\mO_{234} \eqdef \Pr(x^a_2, x^a_3, x^a_4)$:
\begin{align*}
  \mO_{234} = \mH_{234}(O, O, O) \quad\Rightarrow\quad
  \mH_{234} = \mO_{234}(\pinv O, \pinv O, \pinv O).
\end{align*}

@@

Note that in each of the above cases,
  for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  view which is conditionally independent given the other hidden variables in
  the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
A hidden clique $\sC \in \sG$ is said to possess the \textbf{exclusive views property} if for
  every hidden variable $h_i \in \sC$, there is some observed variable,
  $x_{v}$, which is conditionally independent of the rest of the clique
  given $h_i$, and whose conditional moment $\mOpp{v}{i} \eqdef
  \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.

@@

We now show this condition is sufficient to learn the marginal
  distribution of the clique.
Consider any hidden clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$ for which this property holds. Let
  $x_{v_j}$ be any exclusive view for $h_{i_j}$ in $\sC$ and $\sV
  = \{x_{v_1}, \cdots, x_{v_m}\}$ be a set of exclusive views for the clique $\sC$.
We can write down the moments $\mO_\sV$ as follows:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  %&= \sum_{\vh \in \sH}
      %\Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{\vh \in \sH} \Pr(\Sh{\sC}) 
          \Pr(x_{i_1} | h_{v_1}) \cdots \Pr(x_{i_m} | h_{v_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

If each $\mOpp{v_j}{i_j}$ has full column rank, then we can recover the
hidden marginals $Z_\sC$ by inverting:
\begin{align*}
  Z_{\sC} &= \mO_\sV(\pinv{\mOpp{v_1}{i_1}},\cdots,\pinv{\mOpp{v_m}{i_m}}).
\end{align*}
Finally, the conditional probability tables for $\sC$ can easily be obtained from
  $Z_\sC$ via renormalization.
\algorithmref{learnclique} summarizes the procedure, \LearnClique.

\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \sectionref{sampleComplexity}.

@@

  Let $x_v$ be an observed variable with parents $\Pa(x_v)$ and children $\text{Ch}(x_v)$.
  Consider the following transformation.
  Replace $x_v$ with a new hidden variable \hp\ with the same
  parents $\Pa(x_v)$ and children $\text{Ch}(x_v) \union \{x_{v_1}, x_{v_2}, x_{v_3}\}$,
  where $x_{v_1},x_{v_2},x_{v_3}$ are three copies of $x$
  (\figureref{reduction}). 
  The domain of $\hp$ is the same as $x$,
    and $\mOpp{v}{\textrm{new}} = I$.\footnote{
      Though we have assumed that all hidden variables share the
      same domain, $[k]$, our work generalizes to hidden domains of
      different sizes, provided \assumptionref{full-rank-plus} is satisfied.
      }

  Then, there is a one-to-one correspondence between every value of
  $\hp$ and $x_v$. Consequently, for any clique $\sC \contains \hp$, the
  parameters in the original graphical model can be obtained by
  substituting $\hp$ with $x_v$.

@@

We can now use the delta-method to convert the above result into 
the asymptotic variance for the pseudoinverse version of $\LearnClique$:
\begin{lemma}[Asymptotic variance of pseudoinverse estimator for $Z_\sC$]
  \label{lem:mom-variance}  
  Assume $\hat M_{\sV}$ has asymptotic variance $\Sigma_\sV$ defined above.
  Then the asymptotic variance of $\hat Z_\sC$ is:
  \begin{align*}
    \Sigma^{\mom} &= \mOppAlli \Sigma_\sV \mOppAllit.
  \end{align*}
  %where $\pinv{\mOppAll} \eqdef \pinv{\mOpp{v_1}{i_1}} \otimes
  %\cdots \otimes \pinv{\mOpp{v_m}{i_m}}$, a $d^m \times k^m$ matrix.
\end{lemma}
\begin{proof}
For clique $\sC$, recall we have
  $Z_{\sC} = \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m})$.
%Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
%$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
%matrix,
We can rewrite this representation more compactly as: $Z_{\sC} = \pinv{\mOppAll} \mO_\sV$.
By the delta-method \cite{vaart98asymptotic},
% DONE: need to shorten
%we have that the asymptotic variance of $M_\sV$ is:
%\begin{align*}
%  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
%\end{align*}
%where $\Sigma_\sV$ is the variance of the observations. 
we immediately get:
$\sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \mOppAlli \Sigma_\sV \mOppAllit)$.
\end{proof}

>>>>

\paragraph{Estimating $\pi$}
The moments of $x^a_1$, $\mO_1 \eqdef \Pr(x^a_1)$ are directly related to
  $\pi$ by $\mO_1 = O \pi$. 
By \assumptionref{full-rank}, $O$ has full column rank and thus can be
  inverted to recover $\pi$: $\pi = O\pinv  \mO_1$.

\paragraph{Estimating $T$}
Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$, in terms of the hidden marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$:
\begin{align*}
\mO_{12} = O \mH_{12} O^\top \quad\Rightarrow\quad
  \mH_{12} = O\pinv  \mO_{12} O\pinvt .
\end{align*}
$T$ can be recovered from the $\mH_{12}$ by suitable renormalization.

\paragraph{Estimating $V$}
Finally, we can estimate $V$ by renormalizing the hidden marginals
$\mH_{234} \eqdef \Pr(h_2, h_3, h_4)$ from the third-order moments
$\mO_{234} \eqdef \Pr(x^a_2, x^a_3, x^a_4)$:
\begin{align*}
  \mO_{234} = \mH_{234}(O, O, O) \quad\Rightarrow\quad
  \mH_{234} = \mO_{234}(O\pinv , O\pinv , O\pinv ).
\end{align*}

@@

Note that in each of the above cases,
  for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  view which is conditionally independent given the other hidden variables in
  the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
A hidden clique $\sC \in \sG$ is said to possess the \textbf{exclusive views property} if for
  every hidden variable $h_i \in \sC$, there is some observed variable,
  $x_{v}$ which is conditionally independent of the rest of the clique
  given $h_i$, and whose conditional moment $\mOpp{v}{i} \eqdef
  \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.

@@

We now show this condition is sufficient to learn the marginal
  distribution of the clique.
Consider any hidden clique $\sC = \{i_1, \cdots, i_m\}$ for which this property holds. Let
  $x_{v_j}$ be any exclusive view for $h_{i_j}$ in $\sC$ and $\sV
  = \{v_1, \cdots, v_m\}$ be a set of exclusive views for the clique $\sC$.
We can write down the moments $\mO_\sV$ as follows:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  %&= \sum_{\vh \in \sH}
      %\Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{\vh \in \sH} \Pr(\Sh{\sC}) 
          \Pr(x_{v_1} | h_{i_1}) \cdots \Pr(x_{v_m} | h_{i_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

If each $\mOpp{v_j}{i_j}$ has full column rank, then we can recover the
hidden marginals $Z_\sC$ by inverting:
\begin{align*}
  Z_{\sC} &= \mO_\sV(\mOpp{v_1}{i_1}\pinv,\cdots,\mOpp{v_m}{i_m}\pinv).
\end{align*}
Finally, the conditional probability tables for $\sC$ can easily be obtained from
  $Z_\sC$ via renormalization.
\algorithmref{learnclique} summarizes the procedure, \LearnClique.

\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\Sx{\sV}}( \mOpp{v_1}{i_i}\pinv, \cdots, {\mOpp{v_m}{i_m}}\pinv )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \sectionref{sampleComplexity}.

@@

  Let $x_v$ be an observed variable with parents $\Pa(x_v)$ and children $\text{Ch}(x_v)$.
  Consider the following transformation.
  Replace $x_v$ with a new hidden variable \hp\ with the same
  parents $\Pa(x_v)$ and children $\text{Ch}(x_v) \union \{x_{v_1}, x_{v_2}, x_{v_3}\}$,
  where $x_{v_1},x_{v_2},x_{v_3}$ are three copies of $x$
  (\figureref{reduction}). 
  The domain of $\hp$ is the same as $x_v$\footnote{
      Though we have assumed that all hidden variables share the
      same domain, $[k]$, our work generalizes to hidden domains of
      different sizes, provided \assumptionref{full-rank-plus} is satisfied.
      },
    and $\mOpp{v}{\textrm{new}} = I$.
  Then, there is a one-to-one correspondence between every value of
  $\hp$ and $x_v$. Consequently, for any clique $\sC \contains \hp$, the
  parameters in the original graphical model can be obtained by
  substituting $\hp$ with $x_v$.

@@

We can now use the delta-method to convert the above result into 
the asymptotic variance for the pseudoinverse version of $\LearnClique$:
\begin{lemma}[Asymptotic variance of pseudoinverse estimator for $Z_\sC$]
  \label{lem:mom-variance}  
  Assume $\hat M_{\sV}$ has asymptotic variance $\Sigma_\sV$ defined above.
  Then the asymptotic variance of $\hat Z_\sC$ is:
  \begin{align*}
    \Sigma^{\mom} &= \mOppAlli \Sigma_\sV \mOppAllit.
  \end{align*}
  %where ${\mOppAll}\pinv \eqdef {\mOpp{v_1}{i_1}}\pinv \otimes
  %\cdots \otimes {\mOpp{v_m}{i_m}}\pinv$, a $d^m \times k^m$ matrix.
\end{lemma}
\begin{proof}
For clique $\sC$, recall we have
  $Z_{\sC} = \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m})$.
%Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
%$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
%matrix,
We can rewrite this representation more compactly as: $Z_{\sC} = {\mOppAll}\pinv \mO_\sV$.
By the delta-method \cite{vaart98asymptotic},
% DONE: need to shorten
%we have that the asymptotic variance of $M_\sV$ is:
%\begin{align*}
%  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
%\end{align*}
%where $\Sigma_\sV$ is the variance of the observations. 
we immediately get:
$\sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \mOppAlli \Sigma_\sV \mOppAllit)$.
\end{proof}
